Category-based and location-based volitional covert attention affect memory at different timescales

### 2023-03-31 paper/main.tex
Several changes and fixes:   - Added statistical tests for the "place versus face" familiarity differences in both conditions   - Added several citations recommended by reviewers from previous versions of the manuscript

Splitting participants' responses to face versus place images also revealed
that participants often rated attended (and partially attended) place images as
more familiar than attention-matched face images (compare dark versus light
bars in Fig.~\ratingsByCategory). We hypothesized that this might be
explainable by some property of the relevant cognitive processes or by
properties of the stimuli themselves. To help elucidate this distinction, we
examined individual exemplars of the face and place images used in our paradigm
(Fig.~\ref{fig:stimuli}A). By design, the face images had consistent head
sizes, viewing angles, expressions, and so on. In contrast, the place images
varied more substantially across images. For example, some place images
depicted human-made structures; others depicted natural scenes; some depicted
indoor views; others depicted outdoor views; etc. This can also be seen by
averaging the pixel intensity values across images, separately for the face and
place stimuli (Fig.~\ref{fig:stimuli}B). Whereas the average face image retains
many of the landmarks characteristic of most faces (e.g., clearly defined hair,
eyes, nose, mouth, head shape, etc.), the average place image does not show
place-specific features as clearly, aside from a general tendency for the tops
of place images to be lighter than the bottoms of place images. We also
computed the pairwise similarities across images from each stimulus category
(Fig.~\ref{fig:stimuli}C) and found that face images tended to be much more
similar to each other than place images (Fig.~\ref{fig:stimuli}D; $t(115258) =
254.764, p < 0.001$). This analysis indicated to us that our experimental
paradigm was not well-suited to identifying cognitively meaningful stimulus
category differences, since participants' category-specific judgements may be
confounded with within-category image similarity differences.

@@

The notion that location-based attention operates at a faster timescale than
category-based attention is supported by prior work on the deployment of visual
attention~\citep{SotoBlan04, StopEtal07}. Our findings that location-based
attention enhances the processing of attended stimuli whereas category-based
attention suppresses the processing of unattended stimuli is also consistent
with prior work on location-based attention~\citep[e.g.,][]{IttiKoch01} and
category-based attention~\citep[e.g.,][]{MoheEtal14}. Our finding that people
better remember attended stimuli also follows prior work on interactions
between attention and memory~\citep{PallWagn02, ChunTurk07, AlyTurk16,
AlyTurk17, WittEtal18, MorrEtal14, BaleEtal21}. Whereas much of this prior work
focused on elucidating the neural basis of these interactions, our work extends
these prior studies by elucidating the specific and separable behavioral
impacts of location-based attention (enhancement with a fast onset) and
category-based attention (suppression with a slow onset) and on subsequent
recognition memory. Both of these effects persisted throughout the 2~min memory
phases of both conditions. However, future work is needed to elucidate the
longevity of these effects beyond 2 minutes.

Another important area for future study concerns how the flow of information
between different brain structures is modulated according to the focus of
volitional attention-- particularly with respect to pathways from primary
sensory regions (e.g., V1, A1) to regions implicated in encoding ongoing
experiences into memory (e.g, medial temporal lobe structures such as the
hippocampus and entorhinal cortex, prefrontal cortex, etc.). For example,
several studies suggest that attention serves to modulate the \textit{gain} of
specific neural circuits~\citep{TreuTruj99, ChanEtal02, EldaEtal13,
SaliThie00}, effectively facilitating or inhibiting the flow of specific neural
representations~\citep{VartEtal07, LaRoEtal14}. Prior work suggests that
category-based attention may be supported by changes in connectivity with the
thalamus~\citep{Schn11}, whereas location-based attention may be supported by
changes in connectivity with primary visual cortex~\citep{NoudEtal10}. That
category-based and location-based attention are mediated by different brain
structures may explain why these different aspects of attention operate on
different timescales and affect memory differently. A strong test of this
hypothesis would entail directly measuring neural activity patterns as people
modulate their focus of attention (e.g., using functional magnetic resonance
imaging or electroencephalography), and then using neural decoding
approaches~\citep[e.g.,][]{HaxbEtal01, NormEtal06b, MannEtal18, OwenEtal21} to
follow how neural representations of attended (or unattended) stimuli are
transfered from primary sensory regions, to higher order sensory regions, to
memory areas. If the effects of attention on memory are mediated by changes in
network dynamics, the transmission rates of the representations of attended
stimuli from primary sensory regions to memory areas should be facilitated
relative to the transmission rates of unattended stimuli. Further, variability
in these neural changes (e.g., as a participant focuses their attention with
more or less success) should track with behavioral measures of memorability.

>>>>

Splitting participants' responses to face versus place images also revealed
that participants often rated attended (and partially attended) place images as
more familiar than attention-matched face images (compare dark versus light
bars in Fig.~\ratingsByCategory; sustained attention: $t(29) = 1.746, p =
0.091$; variable attention: $t(22) = 2.660, p = 0.014$). We hypothesized that
this might be explainable by either some property of the relevant cognitive
processes or by properties of the stimuli themselves. To help elucidate this
distinction, we examined individual exemplars of the face and place images used
in our paradigm (Fig.~\ref{fig:stimuli}A). By design, the face images had
consistent head sizes, viewing angles, expressions, and so on. In contrast, the
place images varied more substantially across images. For example, some place
images depicted human-made structures; others depicted natural scenes; some
depicted indoor views; others depicted outdoor views; etc. This can also be
seen by averaging the pixel intensity values across images, separately for the
face and place stimuli (Fig.~\ref{fig:stimuli}B). Whereas the average face
image retains many of the landmarks characteristic of most faces (e.g., clearly
defined hair, eyes, nose, mouth, head shape, etc.), the average place image
does not show place-specific features as clearly, aside from a general tendency
for the tops of place images to be lighter than the bottoms of place images. We
also computed the pairwise similarities across images from each stimulus
category (Fig.~\ref{fig:stimuli}C) and found that face images tended to be much
more similar to each other than place images (Fig.~\ref{fig:stimuli}D;
$t(115258) = 254.764, p < 0.001$). This analysis indicated to us that our
experimental paradigm was not well-suited to identifying cognitively meaningful
stimulus category differences, since participants' category-specific judgements
may be confounded with within-category image similarity differences.

@@

The notion that location-based attention operates at a faster timescale than
category-based attention is supported by prior work on the deployment of visual
attention~\citep{SotoBlan04, StopEtal07}. Our findings that location-based
attention enhances the processing of attended stimuli whereas category-based
attention suppresses the processing of unattended stimuli is also consistent
with prior work on location-based attention~\citep[e.g.,][]{IttiKoch01} and
category-based attention~\citep[e.g.,][]{MoheEtal14}. Our finding that people
better remember attended stimuli also follows prior work on interactions
between attention and memory~\citep{PallWagn02, ChunTurk07, AlyTurk16,
AlyTurk17, WittEtal18, MorrEtal14, BaleEtal21, UncaEtal11, TurkEtal13,
LaRoEtal15}. Whereas much of this prior work focused on elucidating the neural
basis of these interactions, our work extends these prior studies by
elucidating the specific and separable behavioral impacts of location-based
attention (enhancement with a fast onset) and category-based attention
(suppression with a slow onset) and on subsequent recognition memory. Both of
these effects persisted throughout the 2~min memory phases of both conditions.
However, future work is needed to elucidate the longevity of these effects
beyond 2 minutes.

Another important area for future study concerns how the flow of information
between different brain structures is modulated according to the focus of
volitional attention-- particularly with respect to pathways from primary
sensory regions (e.g., V1, A1) to regions implicated in encoding ongoing
experiences into memory (e.g, medial temporal lobe structures such as the
hippocampus and entorhinal cortex, prefrontal cortex, etc.). For example,
several studies suggest that attention serves to modulate the \textit{gain} of
specific neural circuits~\citep{TreuTruj99, ChanEtal02, EldaEtal13,
SaliThie00}, effectively facilitating or inhibiting the flow of specific neural
representations~\citep{VartEtal07, LaRoEtal14}. Prior work suggests that
category-based attention may be supported by changes in connectivity with the
thalamus~\citep{Schn11}, whereas location-based attention may be supported by
changes in connectivity with primary visual cortex~\citep{NoudEtal10}. That
category-based and location-based attention are mediated by different brain
structures~\citep[e.g.,][ and others]{GiesEtal03} may explain why these
different aspects of attention operate on different timescales and affect
memory differently. A strong test of this hypothesis would entail directly
measuring neural activity patterns as people modulate their focus of attention
(e.g., using functional magnetic resonance imaging or electroencephalography),
and then using neural decoding approaches~\citep[e.g.,][]{HaxbEtal01,
NormEtal06b, MannEtal18, OwenEtal21} to follow how neural representations of
attended (or unattended) stimuli are transfered from primary sensory regions,
to higher order sensory regions, to memory areas. If the effects of attention
on memory are mediated by changes in network dynamics, the transmission rates
of the representations of attended stimuli from primary sensory regions to
memory areas should be facilitated relative to the transmission rates of
unattended stimuli. Further, variability in these neural changes (e.g., as a
participant focuses their attention with more or less success) should track
with behavioral measures of memorability.
